A Logical Solution of the Linearity Challenge

### 2025-08-02 paper/introduction.tex
++

We value these differences because their combined effect allows us to obtain a
remarkably simple formalization of \Calculus.
%
The fact of working with a calculus based on linear logic prevents \emph{by
construction} the same (sequential) process to own both endpoints of a session.
This situation is clearly undesirable since it does not correspond to a useful
pattern of interaction (every meaningful session requires its endpoints to be
used by parallel processes) and is a potential source of deadlocks.
%
From the standpoint of the formalization, where the representation of channel
names is a primary design choice, it spares us the need to \emph{distinguish}
the two endpoints of a session, \eg by means of polarities~\citep{GayHole05} or
by using different names connected by the same
binder~\citep{Vasconcelos12,GayVasconcelos25}.
%
Also, the fact of working with a calculus with \emph{linear} channels and
explicit continuations spares us the need to \emph{update} the type of channels
in typing contexts. When a channel is used it is effectively consumed, therefore
its type is \emph{removed} from the typing context and the type of the
continuation channel is \emph{added} back to the typing context. As is turns
out, removing and adding types is easier than updating them. Even more so
considering that the type of continuation channels can be conveniently added
\emph{at the beginning} of a typing context, since such channels are fresh by
definition.

The rest of the paper is structured as follows. \Cref{sec:calculus} describes
the syntax and the operational semantics of \Calculus and states the properties
of well-typed processes that we formalize and prove. \Cref{sec:agda} outlines
the essential aspects of the formalization with particular emphasis on the
representation of channel names and typing contexts. \Cref{sec:conclusion}
summarizes our contributions, discusses some extensions and future work. The
formalization has been carried out using Agda and its full code is available
from a public
repository.\footnote{\url{https://github.com/boystrange/LinearityChallenge/tree/main/exp}}

>>>>

We value these differences because their combined effect allows us to obtain a
remarkably simple formalization of \Calculus.
%
The fact of working with a calculus based on linear logic prevents \emph{by
construction} the same (sequential) process to own both endpoints of a session.
This situation is clearly undesirable since it does not correspond to a useful
pattern of interaction (every meaningful session requires its endpoints to be
used by parallel processes) and is a potential source of deadlocks.
%
From the standpoint of the formalization, where the representation of channel
names is a primary design choice, it spares us the need to \emph{distinguish}
the two endpoints of a session, \eg by means of polarities~\citep{GayHole05} or
by using different names connected by the same
binder~\citep{Vasconcelos12,GayVasconcelos25}.
%
Also, the fact of working with a calculus with \emph{linear} channels and
explicit continuations spares us the need to \emph{update} the type of channels
in typing contexts. When a channel is used it is effectively consumed, therefore
its type can be \emph{removed} from the typing context and the type of the
continuation channel is \emph{added} back to the typing context. As is turns
out, removing and adding types is easier than updating them. Even more so
considering that the type of continuation channels can be conveniently added
\emph{at the beginning} of a typing context, since such channels are fresh by
definition.

The rest of the paper is structured as follows. \Cref{sec:calculus} describes
the syntax and the operational semantics of \Calculus and states the properties
of well-typed processes that we formalize and prove. \Cref{sec:agda} outlines
the essential aspects of the formalization with particular emphasis on the
representation of channel names and typing contexts. \Cref{sec:conclusion}
summarizes our contributions, discusses some extensions and future work. The
formalization has been carried out using Agda and its full code is available
from a public\Luca{Related work? Confronto con altre formalizzazioni?}
repository.\footnote{\url{https://github.com/boystrange/LinearityChallenge/tree/main/exp}}
